---
title: "Assignment1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##A. Loading and preprocessing the data
###1.Load the data (i.e. read.csv())
Load ggplot2 and plyr library
```{r}
library(ggplot2)
library(plyr)
```
Read data into a new data frame called data

```{r}
data <- read.csv("activity.csv")
```


###2.Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
#Format the date column into date class
data$date <- as.Date(data$date)
#Sort out the NA steps value for further calculation 
na <- is.na(data$steps)
cleanData <- data[!na,]
```

##B. What is mean total number of steps taken per day?


###1.Calculate the total number of steps taken per day

Calculate total number of steps based on the clean data and store in the variable totalSPD

```{r}
totalSPD <- aggregate(cleanData$steps, by=list(cleanData$date),sum)
#review the newly created dataframe
head(totalSPD)
#rename the column
totalSPD <- rename(totalSPD, c("Group.1"="Date","x"="totalsteps"))
#review once again
head(totalSPD)
```
    
###2.Make a histogram of the total number of steps taken each day
```{r}
#create object g as layout for the histogram
g <- ggplot(totalSPD, aes(x=totalsteps))
#choose color, dimension and create labels for the histpgram
g + geom_histogram(fill="lightblue", binwidth = 900) + labs(title = "Total Daily Steps", x="Total Steps", y="Frequency")
```
    
###3.Calculate and report the mean and median of the total number of steps taken per day
```{r}
#Mean of steps
mean_og <-mean(totalSPD$totalsteps)
#Median of steps
median_og <-median(totalSPD$totalsteps)
```

##C. What is the average daily activity pattern?

###1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
#create a new data frame avgSBI(average steps by interval) based on the cleanData 
avgSBI <- aggregate(cleanData$steps, by=list(cleanData$interval),mean)
#rename the column
avgSBI <- rename(avgSBI, c("Group.1"="interval","x"="averagesteps"))
#review the data frame
head(avgSBI)
#create the new object "p" to make the plot as required
p <- ggplot(avgSBI, aes(x=interval,y=averagesteps), type="l")
# create the time series plot
p + geom_line(color="blue") + labs(title="Average of Steps by Interval", x="Interval", y="Average Steps")
```

###2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# Find the max Average steps value and store it in the new variable "maxSteps"
maxSteps <- avgSBI[which.max(avgSBI$averagesteps),]
# Review the value
maxSteps
```
##D. Imputing missing values

###1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
# Count the total number of missing value by using the variable "na" defined in the A part of the assignment
sum(na)

```
###2.Devise a strategy for filling in all of the missing values in the dataset
```{r}
#The average steps by Interval will be used to replace the na value 
#Create a new variable meanSI to store the average steps by interval based on the original data frame
meanSI <- aggregate(steps~interval,data,FUN=mean)
#Create a new dataframe that merges the original data frame with the means of steps by interval
newData <- merge(x=data,y=meanSI,by="interval")
#Take a sneakpeak at the new dataframe:
str(newData)
#Replace na steps values in newData with average steps by interval, create a new column called steps
newData$steps <- ifelse(is.na(newData$steps.x),newData$steps.y,newData$steps.x)
```

###3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}

head(newData)

```

###4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day 

```{r}
#Total steps by day based on new data frame
totalSPD_new <- aggregate(steps~date,newData,FUN=sum)
##Set up plot to compare old and new data frame
par(mfrow=c(1,2))
hist(totalSPD_new$steps, col="blue", ylim=c(0,35),xlab = "Steps", ylab ="Frequency", 
     main= "Total Numbers of Steps Each Day (updated data frame)", cex.main=0.7)
hist(totalSPD$totalsteps, col="green", ylim=c(0,35),xlab = "Steps", ylab ="Frequency", 
     main= "Total Numbers of Steps Each Day (original data frame)",cex.main=0.7)


```
```{r}
#comparing the old and new mean/median between the original and updated data frame
par(mfrow=c(2,1))
newmean <- mean(totalSPD_new$steps)
newmedian <- median(totalSPD_new$steps)
paste("New mean:  ",round(newmean,2),", ","Original mean  :",round(mean_og,2),"Difference  :", round(newmean-mean_og,2))
paste("New median:  ",round(newmedian,2),", ","Original median  :",round(median_og,2),"Difference  :", round(newmedian-median_og,2))
```

##E. Are there differences in activity patterns between weekdays and weekends?

###1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
#Install packages "chron" to use the function is.weekend
library(chron)
#Create a new variable "day" to sort out Weekday/Weekend in the new data frame
newData$day <- ifelse(is.weekend(newData$date),"Weekend","Weekday")
head(newData)
#Create new table with average steps over interval and day of week
meanSI_new <- aggregate(steps~interval+day,newData,FUN=mean)
head(meanSI_new)


```
###2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
#Create plot to compare activity pattern between weekday and weekend
ggplot(meanSI_new, aes(x=interval, y=steps),type="l") + geom_line(color="blue", size=0.5) +facet_grid(day~.) +labs(x="Interval", y="Average Steps", title = "Activity Pattern")
```